\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Method for a Precision Spectrograph

### 2020-07-16 paper/paper.tex
Pg 2-3 Hogg Edits + Figures

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.  The move towards 0.1 $\mps$ RV precision, however, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to vet for incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using a segmented polynomial in the dispersion direction, tuned to detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

@@

This recommends a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters :  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly likely with the latest generation of \eprv\ instruments with stringent environmental stabilizing, likely reducing the axes along which the instrument can vary.  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

\lz{Hogg: I think it's stuff like this you wanted to audit for consistency with statistical literature?}
In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{calibration space} of wavelength solutions at all moments.  In the statistics literature, this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

@@

\name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, primarily because the need for good wavelength calibration in this field is so great.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph, though the precision of the returned wavelengths will depend on the available calibration sources (more discussion in Section \ref{sec:others} below) .

\section{Method} \label{sec:method}
The \name\ method is designed to take a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional.

>>>>

Traditionally, wavelength solutions are constructed by fitting a polynomial to lines from a calibration source to describe the relationship between wavelength and pixel for each echelle order \citep{butler1996, lovis2007, cersullo2019}.  In this framework, each calibration image is treated independently.  The returned wavelength solutions worked well at the level of 1 $\mps$ precision.  The move towards 0.1 $\mps$ RV precision, however, necessitates higher-fidelity calibration data and wavelength models.  These models need to account for high-order spatial variations that can arise from small imperfections in the optics of an instrument and non-uniformity in detector pixel sizes/spacing.  There has been significant effort in using an entire set of calibration images to identify incongruous ThAr lines \citep{coffinet2019} or obtain high-resolution Fourier transform spectra of reference cells \citep{wang2020}.  It has also been found that using a segmented polynomial in the dispersion direction, tuned to capture detector defects, better describes the wavelength solution than a single, continuous polynomial \citep{milakovic2020}.

@@

This recommends a calibration methodology that is \emph{non-parametric}, or not defined by a prescribed, analytic function described by a finite number of parameters :  If every point in the spectrograph detector is sufficiently surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous ninth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials are at detector edges).

The second simple idea follows from the observation that most physical systems have only a few dominant degrees of freedom, meaning most spectrographs vary along only a small number of axes in ``calibration space'', or the (very high-dimensional) space of all possible wavelength solutions.  This is particularly likely with the latest generation of \eprv\ instruments that are equipped with stringent environmental stabilizing, likely reducing the number and extent of the axes along which the instrument can vary.  That is, spectrographs, especially stabilized ones, should have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the calibration space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the small, accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution at one moment, but also to determine the possible \emph{calibration space} of wavelength solutions at all moments.  In the statistics literature, this concept is often described as \emph{de-noising}:  we can improve calibration by recognizing that every calibration exposure contains information about every other calibration exposure.  Thus every exposure can be improved (i.e., de-noised) with information from every other exposure.

@@

\name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, primarily because the need for good wavelength calibration is so great in this field.  Irregardless, we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph, though the precision of the returned wavelengths will depend on the available calibration sources (more discussion in Section \ref{sec:others} below) .

\section{Method} \label{sec:method}
The \name\ method is designed to take a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas; The wavelength solution should be given enormous flexibility, but it lives in a very low-dimensional calibration space, where the degrees of freedom are set by the limited kinematics of the spectrograph hardware.  \name\ therefore assumes that the space of possible calibration states for an instrument is low-dimensional while assuming very little about the forms of those states.
